---
title: "Report 11.5"
author: "Francesco Lucano"
date: "03/06/2025" 
output:
  html_document:
    toc: true
---
## 1. Informations on site characteristics 
Here we provide information about the characteristics of the 2 study sites,  Lägeren (CH-Lae) and Davos Seehornwald (CH-Dav). The information are draw from Etzold et al., 2011 and Zielis et al., 2014. 

**Comparative characteristics of CH-Lae (Lägeren) and CH-Dav (Davos Seehornwald) forest sites**

---

**Location**

* CH-Lae: Located in the Swiss Jura mountain range, representative of the montane zone.
* CH-Dav: Situated in the Eastern Swiss Alps, representative of the subalpine zone.

---

**Elevation**

* CH-Lae: 682 
* CH-Dav: 1,639 

---

**Climate (mean annual values)**

**Air Temperature (20-year average 1989-2009)**

* CH-Lae: 7.4 °C 
* CH-Dav: 3.4 °C 

**Precipitation (20-year average 1989-2009)**

* CH-Lae: 1000 mm 
* CH-Dav: 1000 mm 

---

**Vegetation Type & Dominant Species**

* CH-Lae: A mixed deciduous-dominated forest with a seasonal leaf course. Key species include *Fagus sylvatica L.* (European beech), *Picea abies (L.) Karst* (Norway spruce), *Fraxinus excelsior L.* (ash), and *Acer pseudoplatanus L.*.
* CH-Dav: A coniferous forest, specifically a subalpine spruce forest, it's an evergreen site. Highly dominated by *Picea abies (L.) Karst* (Norway spruce).

---

**Maximum Leaf Area Index (LAI)**

* CH-Lae: 1.7-5.5 m$^2$ m$^{-2}$ (reflecting seasonal variation).
* CH-Dav: 3.9 m$^2$ m$^{-2}$ (more stable year-round).

---

**Summary**

In essence, the two sites show marked difference that will affect Gross Primary Production (GPP) and they will works well to assess model perfomance.

CH-Dav, represents a high-altitude subalpine environment characterized by a cold climate and snow precipitations that influence the GPP. Conversely, CH-Lae, a lower-altitude site, experiences a milder temperate climate. This altitudinal and climatic disparity directly translates into divergent growing season lengths and conditions.

This translates into very different vegetation types. CH-Dav is an established evergreen coniferous forest, predominantly composed of spruce, which sustains its canopy throughout the year. Conversely, CH-Lae is a younger mixed forest, prominently composed of deciduous trees, notably beech. This structural difference results in distinct seasonal changes in leaf area and overall canopy coverage at CH-Lae.


## 2. KNN model training

### 2.1. Data preparation
```{r setup, warning=FALSE, message=FALSE}
library(readr)     
library(dplyr)     
library(lubridate) 
library(ggplot2)   
library(rsample)   
library(recipes)   
library(caret)     
library(here)
library(yardstick)
library(tidyr)

# Load CH-Dav data
flux_davos_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Dav_FLUXNET2015_FULLSET_DD_1997-2014_1-3.csv"),
  delim = ","
)

# Load CH-Lae data
flux_laegeren_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Lae_FLUXNET2015_FULLSET_DD_2004-2014_1-4.csv"),
  delim = ","
)
```

```{r message=FALSE, warning=FALSE}
# We define a function in order to clean both dataset
clean_flux_data <- function(raw_df) { 
  df_cleaned <- raw_df %>%
    # We select only the columns we need
    select(
      TIMESTAMP, GPP_NT_VUT_REF, NEE_VUT_REF_QC,
      TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F
    ) %>%
    # We rename GPP column for simplicity
    rename(GPP = GPP_NT_VUT_REF) %>%
    # We convert TIMESTAMP to date object after making it a character 
    mutate(TIMESTAMP = as.Date(as.character(TIMESTAMP), format = "%Y%m%d"),
           # We replace -9999 values with NA
           across(where(is.numeric), ~na_if(.x, -9999))) %>%
    # We filter rows only to keep good quality data (value 1)
    filter(NEE_VUT_REF_QC == 1) %>% 
    # We remove rows where any of the remaining columns have NA values
    select(TIMESTAMP, GPP, TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F) %>%
    filter(complete.cases(.))
  return(df_cleaned)
}

# Then we apply it:
flux_davos_cleaned <- clean_flux_data(flux_davos_raw)
flux_laegeren_cleaned <- clean_flux_data(flux_laegeren_raw)
```

```{r, warning=FALSE, message=FALSE}
# We add a site_id column to each cleaned dataset
flux_davos_cleaned_labelled <- flux_davos_cleaned %>%
  mutate(site_id = "CH-Dav")

flux_laegeren_cleaned_labelled <- flux_laegeren_cleaned %>%
  mutate(site_id = "CH-Lae")

# We combine the two datasets
flux_combined_cleaned <- bind_rows(flux_davos_cleaned_labelled, flux_laegeren_cleaned_labelled)

# We convert site_id to a factor
flux_combined_cleaned <- flux_combined_cleaned %>%
  mutate(site_id = as.factor(site_id))

# Summary
print("Combined cleaned data head:")
print(head(flux_combined_cleaned))
print(dim(flux_combined_cleaned))
print("Summary of combined data:")
print(summary(flux_combined_cleaned))
print("Counts per site:")
print(table(flux_combined_cleaned$site_id))
```


The goal of this step is to train three distinct, optimally tuned K-Nearest Neighbors (KNN) models:

    - Model D: Trained using only the Davos data (flux_davos_cleaned_labelled).
    - Model L: Trained using only the Laegern data (flux_laegeren_cleaned_labelled).
    - Model C: Trained using the Combined data (flux_combined_cleaned).


### 2.2. Data splitting 
```{r, warning=FALSE, message=FALSE}
# We define the data splitting function to use for each dataset 
split_data_stratified <- function(input_df, 
                                  split_proportion = 0.80,   # (80% training, 30% testing)
                                  stratification_variable) { 

  split_object <- initial_split(
    data = input_df, 
    prop = split_proportion, 
    strata = {{ stratification_variable }} 
  )
  
  # Extract the training data and testing data
  training_set <- training(split_object)
  testing_set <- testing(split_object)
  
  # Return a list containing the training and testing sets
  return(list(train = training_set, test = testing_set))
}
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 
davos_splits <- split_data_stratified(
  input_df = flux_davos_cleaned_labelled, 
  split_proportion = 0.80, 
  stratification_variable = VPD_F # We stratify by vapour pressure deficit to ensure that the distribution is similar in both the training set and the testing set
)
davos_train <- davos_splits$train
davos_test <- davos_splits$test

# Check dimensions
print("Davos Train Dimensions:")
print(dim(davos_train))
print("Davos Test Dimensions:")
print(dim(davos_test))

set.seed(60249) 
laegeren_splits <- split_data_stratified(
  input_df = flux_laegeren_cleaned_labelled, 
  split_proportion = 0.80, 
  stratification_variable = VPD_F # We stratify by vapour pressure deficit to ensure that the distribution is similar in both the training set and the testing set
)
laegeren_train <- laegeren_splits$train
laegeren_test <- laegeren_splits$test

# Check dimensions
print("Laegern Train Dimensions:")
print(dim(laegeren_train))
print("Laegern Test Dimensions:")
print(dim(laegeren_test))

set.seed(60249) 
combined_splits <- split_data_stratified(
  input_df = flux_combined_cleaned, 
  split_proportion = 0.80, 
  stratification_variable = site_id # We stratify by site_id for combined data
)
combined_train <- combined_splits$train
combined_test <- combined_splits$test

# Check dimensions
print("Combined Train Dimensions:")
print(dim(combined_train))
print("Combined Test Dimensions:")
print(dim(combined_test))
print("Site distribution in Combined Train:")
print(table(combined_train$site_id))
print("Site distribution in Combined Test:")
print(table(combined_test$site_id))
```


### 2.3. Preprocessing
```{r, warning=FALSE, message=FALSE}
# We define a function for preprocessing each training dataset
preprocessing <- function(input_df) {
# We define the recipe and we select the numeric predictors 
  gpp_recipe <- recipe(GPP ~ TA_F + SW_IN_F + LW_IN_F + VPD_F + P_F + WS_F, data = input_df) %>%
  # We apply Yeo-Johnson transformation to all numeric predictors (makes the distributions of the variables more normal-like). We use it over the Box-Cox because because Box-Cox requires strictly positive (> 0) values, and our data includes non-positive values in TA_F (negative) and P_F (zero)
    step_YeoJohnson(all_predictors()) %>%
  # We center all numeric predictors (subtract mean)
    step_center(all_predictors()) %>%
  # We scale all numeric predictors (divide by standard deviation)
    step_scale(all_predictors())
  
    return(gpp_recipe) # Explicitly return the recipe object
}
```

```{r, warning=FALSE, message=FALSE}
gpp_recipe_davos <- preprocessing(input_df = davos_train)
print(gpp_recipe_davos)

gpp_recipe_laegeren <- preprocessing(input_df = laegeren_train)
print(gpp_recipe_laegeren)

gpp_recipe_combined <- preprocessing(input_df = combined_train)
print(gpp_recipe_combined)
```


### 2.4. Hyperparameter tuning for the optimal k
```{r, warning=FALSE, message=FALSE}
# we set up training control for 10-fold cross-validation
train_control_cv <- trainControl(
  method = "cv",      
  number = 10        
)

# We define the range of k values we want to test
k_tuning_grid <- data.frame(k = 1:30)
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_davos_tuned <- train(
  gpp_recipe_davos,        
  data = davos_train,      
  method = "knn",
  trControl = train_control_cv, 
  tuneGrid = k_tuning_grid,  
  metric = "MAE"             
  
)

print("Davos KNN model tuning summary:")
print(model_davos_tuned)
print(paste("Optimal k for Davos-only model (based on MAE):", model_davos_tuned$bestTune$k))
plot(model_davos_tuned, main = "CV performance for Davos KNN Model")
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_laegeren_tuned <- train(
  gpp_recipe_laegeren,     
  data = laegeren_train,   
  method = "knn",
  trControl = train_control_cv,
  tuneGrid = k_tuning_grid,
  metric = "MAE"
)

print("Laegern KNN model tuning summary:")
print(model_laegeren_tuned)
print(paste("Optimal k for Laegern-only model (based on MAE):", model_laegeren_tuned$bestTune$k))
plot(model_laegeren_tuned, main = "CV performance for Laegern KNN Model")
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_combined_tuned <- train(
  gpp_recipe_combined,    
  data = combined_train,  
  method = "knn",
  trControl = train_control_cv,
  tuneGrid = k_tuning_grid,
  metric = "MAE"
)

print("Combined KNN model tuning summary:")
print(model_combined_tuned)
print(paste("Optimal k for Combined data model (based on MAE):", model_combined_tuned$bestTune$k))
plot(model_combined_tuned, main = "CV performance for combined KNN Model")
```


## 3. Model evaluation on test set
```{r, warning=FALSE, message=FALSE}
# We define a function to calculate metrics and create plots
eval_model <- function(model, test_data, model_name) {

  # We make predictions on the test data
  predictions <- predict(model, newdata = test_data)

  # We create a data holding observed and predicted values
  eval_data <- data.frame(
    Observed = test_data$GPP,
    Predicted = predictions
  )

  # We calculate performance metrics using the built-in functions from the yardstick package
  rmse_calc <- yardstick::rmse_vec(eval_data$Observed, eval_data$Predicted)
  rsquared_calc <- yardstick::rsq_vec(eval_data$Observed, eval_data$Predicted)

  # Create scatter plot
  p <- ggplot(eval_data, aes(x = Observed, y = Predicted)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", linewidth = 1) + # Use linewidth
    labs(
      title = paste(model_name, ": Observed vs. Predicted GPP (test set)"),
      subtitle = paste("RMSE:", round(rmse_calc, 3), " R²:", round(rsquared_calc, 3)), # MAE removed
      x = "Observed GPP",
      y = "Predicted GPP"
    ) +
    theme_minimal() +
    coord_fixed(ratio = 1) # This ensures axes have the same scale for 1:1 comparison
  
  print(p) # Print the plot during the function call
  
  # We create a data frame of the results 
  metrics_df <- data.frame(
    Evaluation_Scenario = model_name, # This was 'Evaluation_against_test_set = test_set_descriptor'
    R_squared = round(rsquared_calc, 3),
    RMSE = round(rmse_calc, 3)
  )
  
  return(metrics_df)
}
```

```{r, warning=FALSE, message=FALSE}
cat("Evaluating model trained on Dacos data\n")
davos_on_davos_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = davos_test, 
    model_name = "Davos-trained on Davos test" 
)
davos_on_laegeren_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = laegeren_test, 
    model_name = "Davos-trained on Laegern test"
)
davos_on_combined_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = combined_test, 
    model_name = "Davos-trained on combined test"
)

# Create Table 1
table_model_davos_results <- dplyr::bind_rows(
  davos_on_davos_metrics, 
  davos_on_laegeren_metrics, 
  davos_on_combined_metrics
)
knitr::kable(table_model_davos_results, 
             caption = "Performance of KNN model trained on Davos data only.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))


cat("Evaluating model trained on Laegern data\n")
laegeren_on_laegeren_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = laegeren_test, 
  model_name = "Laegern-trained on Laegern test" 
)
laegeren_on_davos_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = davos_test, 
  model_name = "Laegern-trained on Davos test"
)
laegeren_on_combined_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = combined_test, 
  model_name = "Laegern-trained on combined test"
)

# Create Table 2
table_model_laegeren_results <- dplyr::bind_rows(
  laegeren_on_laegeren_metrics, 
  laegeren_on_davos_metrics, 
  laegeren_on_combined_metrics
)
knitr::kable(table_model_laegeren_results, 
             caption = "Performance of KNN model trained on Laegern data only.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))


cat("Evaluating model trained on combined data\n")
combined_on_davos_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = davos_test, 
  model_name = "Combined-trained on Davos test"
)
combined_on_laegeren_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = laegeren_test, 
  model_name = "Combined-trained on Laegern test"
)
combined_on_combined_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = combined_test, 
  model_name = "Combined-trained on combined test"
)

# Create Table 3
table_model_combined_results <- dplyr::bind_rows(
  combined_on_davos_metrics, 
  combined_on_laegeren_metrics, 
  combined_on_combined_metrics
)
knitr::kable(table_model_combined_results, 
             caption = "Performance of KNN model trained on combined (Davos & Laegern) data.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))
```