---
title: "Report 11.5"
author: "Francesco Lucano"
date: "03/06/2025" 
output:
  html_document:
    toc: true
---
## 1. Informations on site characteristics 
Here we provide information about the characteristics of the 2 study sites,  Lägeren (CH-Lae) and Davos Seehornwald (CH-Dav). The information are draw from Etzold et al., 2011 and Zielis et al., 2014. 

**Comparative characteristics of CH-Lae (Lägeren) and CH-Dav (Davos Seehornwald) forest sites**

---

**Location**

* CH-Lae: Located in the Swiss Jura mountain range, representative of mid-altitude mountain environment.
* CH-Dav: Situated in the Eastern Swiss Alps, representative of the subalpine zone.

---

**Elevation**

* CH-Lae: 682 
* CH-Dav: 1,639 

---

**Climate (mean annual values)**

**Air Temperature (20-year average 1989-2009)**

* CH-Lae: 7.4 °C 
* CH-Dav: 3.4 °C 

**Precipitation (20-year average 1989-2009)**

* CH-Lae: 1000 mm 
* CH-Dav: 1000 mm 

---

**Vegetation Type & Dominant Species**

* CH-Lae: A mixed deciduous-dominated forest with a seasonal leaf course. Key species include *Fagus sylvatica L.* (European beech), *Picea abies (L.) Karst* (Norway spruce), *Fraxinus excelsior L.* (ash), and *Acer pseudoplatanus L.*.
* CH-Dav: A coniferous forest, specifically a subalpine spruce forest, it's an evergreen site. Highly dominated by *Picea abies (L.) Karst* (Norway spruce).

---

**Maximum Leaf Area Index (LAI)**

* CH-Lae: 1.7-5.5 m$^2$ m$^{-2}$ (reflecting seasonal variation).
* CH-Dav: 3.9 m$^2$ m$^{-2}$ (more stable year-round).

---

**Summary**

In essence, the two sites show marked difference that will affect Gross Primary Production (GPP) and they will works well to assess model perfomance.

CH-Dav, represents a high-altitude subalpine environment characterized by a cold climate and snow precipitations that influence the GPP. Conversely, CH-Lae, a lower-altitude site, experiences a milder temperate climate. This altitudinal and climatic disparity directly translates into divergent growing season lengths and conditions.

This translates into very different vegetation types. CH-Dav is an established evergreen coniferous forest, predominantly composed of spruce, which sustains its canopy throughout the year. Conversely, CH-Lae is a younger mixed forest, prominently composed of deciduous trees, notably beech. This structural difference results in distinct seasonal changes in leaf area and overall canopy coverage at CH-Lae.


## 2. KNN model training

### 2.1. Data preparation
```{r setup, warning=FALSE, message=FALSE}
library(readr)     
library(dplyr)     
library(lubridate) 
library(ggplot2)   
library(rsample)   
library(recipes)   
library(caret)     
library(here)
library(yardstick)
library(tidyr)

# Load CH-Dav data
flux_davos_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Dav_FLUXNET2015_FULLSET_DD_1997-2014_1-3.csv"),
  delim = ","
)

# Load CH-Lae data
flux_laegeren_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Lae_FLUXNET2015_FULLSET_DD_2004-2014_1-4.csv"),
  delim = ","
)
```

```{r message=FALSE, warning=FALSE}
# We define a function in order to clean both dataset
clean_flux_data <- function(raw_df) { 
  df_cleaned <- raw_df %>%
    # We select only the columns we need
    select(
      TIMESTAMP, GPP_NT_VUT_REF, NEE_VUT_REF_QC,
      TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F
    ) %>%
    # We rename GPP column for simplicity
    rename(GPP = GPP_NT_VUT_REF) %>%
    # We convert TIMESTAMP to date object after making it a character 
    mutate(TIMESTAMP = as.Date(as.character(TIMESTAMP), format = "%Y%m%d"),
           # We replace -9999 values with NA
           across(where(is.numeric), ~na_if(.x, -9999))) %>%
    # We filter rows only to keep good quality data (value 1)
    filter(NEE_VUT_REF_QC == 1) %>% 
    # We remove rows where any of the remaining columns have NA values
    select(TIMESTAMP, GPP, TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F) %>%
    filter(complete.cases(.))
  return(df_cleaned)
}

# Then we apply it:
flux_davos_cleaned <- clean_flux_data(flux_davos_raw)
flux_laegeren_cleaned <- clean_flux_data(flux_laegeren_raw)
```

```{r, warning=FALSE, message=FALSE}
# We add a site_id column to each cleaned dataset
flux_davos_cleaned_labelled <- flux_davos_cleaned %>%
  mutate(site_id = "CH-Dav")

flux_laegeren_cleaned_labelled <- flux_laegeren_cleaned %>%
  mutate(site_id = "CH-Lae")

# We combine the two datasets
flux_combined_cleaned <- bind_rows(flux_davos_cleaned_labelled, flux_laegeren_cleaned_labelled)

# We convert site_id to a factor
flux_combined_cleaned <- flux_combined_cleaned %>%
  mutate(site_id = as.factor(site_id))

# Summary
print("Combined cleaned data head:")
print(head(flux_combined_cleaned))
print(dim(flux_combined_cleaned))
print("Summary of combined data:")
print(summary(flux_combined_cleaned))
print("Counts per site:")
print(table(flux_combined_cleaned$site_id))
```


The goal of this step is to train three distinct, optimally tuned K-Nearest Neighbors (KNN) models:

    - Model D: Trained using only the Davos data (flux_davos_cleaned_labelled).
    - Model L: Trained using only the Laegern data (flux_laegeren_cleaned_labelled).
    - Model C: Trained using the Combined data (flux_combined_cleaned).


### 2.2. Data splitting 
```{r, warning=FALSE, message=FALSE}
# We define the data splitting function to use for each dataset 
split_data_stratified <- function(input_df, 
                                  split_proportion = 0.80,   # (80% training, 30% testing)
                                  stratification_variable) { 

  split_object <- initial_split(
    data = input_df, 
    prop = split_proportion, 
    strata = {{ stratification_variable }} 
  )
  
  # Extract the training data and testing data
  training_set <- training(split_object)
  testing_set <- testing(split_object)
  
  # Return a list containing the training and testing sets
  return(list(train = training_set, test = testing_set))
}
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 
davos_splits <- split_data_stratified(
  input_df = flux_davos_cleaned_labelled, 
  split_proportion = 0.80, 
  stratification_variable = VPD_F # We stratify by vapour pressure deficit to ensure that the distribution is similar in both the training set and the testing set
)
davos_train <- davos_splits$train
davos_test <- davos_splits$test

# Check dimensions
print("Davos Train Dimensions:")
print(dim(davos_train))
print("Davos Test Dimensions:")
print(dim(davos_test))

set.seed(60249) 
laegeren_splits <- split_data_stratified(
  input_df = flux_laegeren_cleaned_labelled, 
  split_proportion = 0.80, 
  stratification_variable = VPD_F # We stratify by vapour pressure deficit to ensure that the distribution is similar in both the training set and the testing set
)
laegeren_train <- laegeren_splits$train
laegeren_test <- laegeren_splits$test

# Check dimensions
print("Laegern Train Dimensions:")
print(dim(laegeren_train))
print("Laegern Test Dimensions:")
print(dim(laegeren_test))

set.seed(60249) 
combined_splits <- split_data_stratified(
  input_df = flux_combined_cleaned, 
  split_proportion = 0.80, 
  stratification_variable = site_id # We stratify by site_id for combined data
)
combined_train <- combined_splits$train
combined_test <- combined_splits$test

# Check dimensions
print("Combined Train Dimensions:")
print(dim(combined_train))
print("Combined Test Dimensions:")
print(dim(combined_test))
print("Site distribution in Combined Train:")
print(table(combined_train$site_id))
print("Site distribution in Combined Test:")
print(table(combined_test$site_id))
```


### 2.3. Preprocessing
```{r, warning=FALSE, message=FALSE}
# We define a function for preprocessing each training dataset
preprocessing <- function(input_df) {
# We define the recipe and we select the numeric predictors 
  gpp_recipe <- recipe(GPP ~ TA_F + SW_IN_F + LW_IN_F + VPD_F + P_F + WS_F, data = input_df) %>%
  # We apply Yeo-Johnson transformation to all numeric predictors (makes the distributions of the variables more normal-like). We use it over the Box-Cox because because Box-Cox requires strictly positive (> 0) values, and our data includes non-positive values in TA_F (negative) and P_F (zero)
    step_YeoJohnson(all_predictors()) %>%
  # We center all numeric predictors (subtract mean)
    step_center(all_predictors()) %>%
  # We scale all numeric predictors (divide by standard deviation)
    step_scale(all_predictors())
  
    return(gpp_recipe) # Explicitly return the recipe object
}
```

```{r, warning=FALSE, message=FALSE}
gpp_recipe_davos <- preprocessing(input_df = davos_train)
print(gpp_recipe_davos)

gpp_recipe_laegeren <- preprocessing(input_df = laegeren_train)
print(gpp_recipe_laegeren)

gpp_recipe_combined <- preprocessing(input_df = combined_train)
print(gpp_recipe_combined)
```


### 2.4. Hyperparameter tuning for the optimal k
```{r, warning=FALSE, message=FALSE}
# we set up training control for 10-fold cross-validation
train_control_cv <- trainControl(
  method = "cv",      
  number = 10        
)

# We define the range of k values we want to test
k_tuning_grid <- data.frame(k = 1:30)
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_davos_tuned <- train(
  gpp_recipe_davos,        
  data = davos_train,      
  method = "knn",
  trControl = train_control_cv, 
  tuneGrid = k_tuning_grid,  
  metric = "MAE"             
  
)

print("Davos KNN model tuning summary:")
print(model_davos_tuned)
print(paste("Optimal k for Davos-only model (based on MAE):", model_davos_tuned$bestTune$k))
plot(model_davos_tuned, main = "CV performance for Davos KNN Model")
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_laegeren_tuned <- train(
  gpp_recipe_laegeren,     
  data = laegeren_train,   
  method = "knn",
  trControl = train_control_cv,
  tuneGrid = k_tuning_grid,
  metric = "MAE"
)

print("Laegern KNN model tuning summary:")
print(model_laegeren_tuned)
print(paste("Optimal k for Laegern-only model (based on MAE):", model_laegeren_tuned$bestTune$k))
plot(model_laegeren_tuned, main = "CV performance for Laegern KNN Model")
```

```{r, warning=FALSE, message=FALSE}
set.seed(60249) 

model_combined_tuned <- train(
  gpp_recipe_combined,    
  data = combined_train,  
  method = "knn",
  trControl = train_control_cv,
  tuneGrid = k_tuning_grid,
  metric = "MAE"
)

print("Combined KNN model tuning summary:")
print(model_combined_tuned)
print(paste("Optimal k for Combined data model (based on MAE):", model_combined_tuned$bestTune$k))
plot(model_combined_tuned, main = "CV performance for combined KNN Model")
```


## 3. Model evaluation on test set
```{r, warning=FALSE, message=FALSE}
# We define a function to calculate metrics and create plots
eval_model <- function(model, test_data, model_name) {

  # We make predictions on the test data
  predictions <- predict(model, newdata = test_data)

  # We create a data holding observed and predicted values
  eval_data <- data.frame(
    Observed = test_data$GPP,
    Predicted = predictions
  )

  # We calculate performance metrics using the built-in functions from the yardstick package
  rmse_calc <- yardstick::rmse_vec(eval_data$Observed, eval_data$Predicted)
  rsquared_calc <- yardstick::rsq_vec(eval_data$Observed, eval_data$Predicted)

  # Create scatter plot
  p <- ggplot(eval_data, aes(x = Observed, y = Predicted)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", linewidth = 1) + # Use linewidth
    labs(
      title = paste(model_name, ": Observed vs. Predicted GPP (test set)"),
      subtitle = paste("RMSE:", round(rmse_calc, 3), " R²:", round(rsquared_calc, 3)), # MAE removed
      x = "Observed GPP",
      y = "Predicted GPP"
    ) +
    theme_minimal() +
    coord_fixed(ratio = 1) # This ensures axes have the same scale for 1:1 comparison
  
  print(p) # To print the plot 
  
  # We create a data frame of the results 
  metrics_df <- data.frame(
    Evaluation_Scenario = model_name, 
    R_squared = round(rsquared_calc, 3),
    RMSE = round(rmse_calc, 3)
  )
  
  return(metrics_df)
}
```

```{r, warning=FALSE, message=FALSE}
cat("Evaluating model trained on Dacos data\n")
davos_on_davos_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = davos_test, 
    model_name = "Davos-trained on Davos test" 
)
davos_on_laegeren_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = laegeren_test, 
    model_name = "Davos-trained on Laegern test"
)
davos_on_combined_metrics <- eval_model(
    model = model_davos_tuned, 
    test_data = combined_test, 
    model_name = "Davos-trained on combined test"
)

# Create Table 1
table_model_davos_results <- dplyr::bind_rows(
  davos_on_davos_metrics, 
  davos_on_laegeren_metrics, 
  davos_on_combined_metrics
)
knitr::kable(table_model_davos_results, 
             caption = "Performance of KNN model trained on Davos data only.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))


cat("Evaluating model trained on Laegern data\n")
laegeren_on_laegeren_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = laegeren_test, 
  model_name = "Laegern-trained on Laegern test" 
)
laegeren_on_davos_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = davos_test, 
  model_name = "Laegern-trained on Davos test"
)
laegeren_on_combined_metrics <- eval_model(
  model = model_laegeren_tuned, 
  test_data = combined_test, 
  model_name = "Laegern-trained on combined test"
)

# Create Table 2
table_model_laegeren_results <- dplyr::bind_rows(
  laegeren_on_laegeren_metrics, 
  laegeren_on_davos_metrics, 
  laegeren_on_combined_metrics
)
knitr::kable(table_model_laegeren_results, 
             caption = "Performance of KNN model trained on Laegern data only.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))


cat("Evaluating model trained on combined data\n")
combined_on_davos_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = davos_test, 
  model_name = "Combined-trained on Davos test"
)
combined_on_laegeren_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = laegeren_test, 
  model_name = "Combined-trained on Laegern test"
)
combined_on_combined_metrics <- eval_model(
  model = model_combined_tuned, 
  test_data = combined_test, 
  model_name = "Combined-trained on combined test"
)

# Create Table 3
table_model_combined_results <- dplyr::bind_rows(
  combined_on_davos_metrics, 
  combined_on_laegeren_metrics, 
  combined_on_combined_metrics
)
knitr::kable(table_model_combined_results, 
             caption = "Performance of KNN model trained on combined (Davos & Laegern) data.",
             col.names = c("Evaluation scenario", "R²", "RMSE"))
```


## 4. Discussion
When evaluating the performance of models trained and tested on data from the same site, we can observe that the model trained on Davos data achieves a higher degree of accuracy for its own environment than the Laegern model does for its environment. The Davos model explains approximately 75% of the variance ($R^2$ = 0.745) on unseen Davos data , whereas the Laegern model explains about 62% of the variance ($R^2$ = 0.621) on its respective test set. This suggests that the GPP at the Davos site may be more consistently predicted by the chosen meteorological variables.

However, when assessing the models performance on the unseen site, the Davos-trained model performs significantly worse when transferred. When applied to the Laegern test set, its explanatory power drops considerably ($R^2$ from 0.745 to 0.536) and its prediction error increases substantially (RMSE from 1.411 to 2.818). Curiously, the model trained on Laegern data does not show a similar degradation when predicting for Davos. Its $R^2$ remains almost the same (0.621 vs 0.612), and its RMSE actually improves when predicting for Davos compared to its own test set (2.205 vs 2.473). This could be due to the much greater environmental and vegetative variety at the Laegern site, which may lead to a model that has learned more robust and general patterns.

Finally, the results for the model trained on the combined dataset show an interesting behavior, as it outperforms both specialized models on their home test sets. For example, the combined models prediction for Davos data ($R^2$ = 0.761)  is slightly more accurate than that of the specialized Davos model ($R^2$ = 0.745). Similarly, its prediction for Laegern data ($R^2$ = 0.646)  is also better than that of the specialized Laegern model ($R^2$ = 0.621). This suggests that training on a more diverse dataset from both sites created a more robust and generalized model. For context, the specialized models performed moderately on the combined test set, with the Davos-trained model achieving an $R^2$ of 0.585  and the Laegern-trained model an $R^2$ of 0.587.


### 4.1. Time series analysis
```{r, warning=FALSE, message=FALSE}
# Time series visualization of the models for the Davos test set
# We create and reshape the data for the Davos test set 
ts_davos_plot_data <- davos_test %>%
  select(TIMESTAMP, GPP) %>%
  mutate(
    Predicted_by_Davos_Model = predict(model_davos_tuned, newdata = davos_test),
    Predicted_by_Laegern_Model = predict(model_laegeren_tuned, newdata = davos_test),
    Predicted_by_Combined_Model = predict(model_combined_tuned, newdata = davos_test)
  ) %>%
  rename(Observed_GPP = GPP) 

ts_davos_plot_long <- ts_davos_plot_data %>%
  pivot_longer(
    cols = c(Observed_GPP, Predicted_by_Davos_Model, Predicted_by_Laegern_Model, Predicted_by_Combined_Model),
    names_to = "Source",
    values_to = "GPP_Value"
  )

# We filter from 2010 to 2014 for a clearer visualization 
ts_davos_plot_subset <- ts_davos_plot_long %>%
  filter(year(TIMESTAMP) >= 2010 & year(TIMESTAMP) < 2014)

# Create the plot using the filtered data
plot_davos_comparison_subset <- ggplot(ts_davos_plot_subset, aes(x = TIMESTAMP, y = GPP_Value, color = Source)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    title = "Model predictions vs. observed GPP on the Davos test set",
    subtitle = "Time series evaluation (2010-2014)",
    x = "Date",
    y = expression("GPP (gC m"^-2*" day"^-1*")"),
    color = "Data source:"
  ) +
  theme_minimal()

print(plot_davos_comparison_subset)


# We do the same for the Laegern test set
ts_laegeren_plot_data <- laegeren_test %>%
  select(TIMESTAMP, GPP) %>%
  mutate(
    Predicted_by_Davos_Model = predict(model_davos_tuned, newdata = laegeren_test),
    Predicted_by_Laegern_Model = predict(model_laegeren_tuned, newdata = laegeren_test),
    Predicted_by_Combined_Model = predict(model_combined_tuned, newdata = laegeren_test)
  ) %>%
  rename(Observed_GPP = GPP)

ts_laegeren_plot_long <- ts_laegeren_plot_data %>%
  pivot_longer(
    cols = c(Observed_GPP, Predicted_by_Davos_Model, Predicted_by_Laegern_Model, Predicted_by_Combined_Model),
    names_to = "Source",
    values_to = "GPP_Value"
  )

ts_laegeren_plot_subset <- ts_laegeren_plot_long %>%
  filter(year(TIMESTAMP) >= 2010 & year(TIMESTAMP) < 2015)


plot_laegeren_comparison_subset <- ggplot(ts_laegeren_plot_subset, aes(x = TIMESTAMP, y = GPP_Value, color = Source)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  labs(
    title = "Model predictions vs. observed GPP on the Laegeren test set",
    subtitle = "Time series evaluation (2010-2014)",
    x = "Date",
    y = expression("GPP (gC m"^-2*" day"^-1*")"),
    color = "Data source:"
  ) +
  theme_minimal()

print(plot_laegeren_comparison_subset)


# And for the combined test set
ts_combined_plot_data <- combined_test %>%
  select(TIMESTAMP, GPP, site_id) %>%
  mutate(
    Predicted_by_Davos_Model = predict(model_davos_tuned, newdata = combined_test),
    Predicted_by_Laegern_Model = predict(model_laegeren_tuned, newdata = combined_test),
    Predicted_by_Combined_Model = predict(model_combined_tuned, newdata = combined_test)
  ) %>%
  rename(Observed_GPP = GPP)

ts_combined_plot_long <- ts_combined_plot_data %>%
  pivot_longer(
    cols = c(Observed_GPP, Predicted_by_Davos_Model, Predicted_by_Laegern_Model, Predicted_by_Combined_Model),
    names_to = "Source",
    values_to = "GPP_Value"
  )

ts_combined_plot_subset <- ts_combined_plot_long %>%
  filter(year(TIMESTAMP) >= 2010 & year(TIMESTAMP) < 2015)

plot_combined_comparison_subset <- ggplot(ts_combined_plot_subset, aes(x = TIMESTAMP, y = GPP_Value, color = Source)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  facet_wrap(~ site_id, scales = "free_y", ncol = 1) + # Separate panels for CH-Dav and CH-Lae data
  labs(
    title = "Model predictions vs. observed GPP on the combined test set",
    subtitle = "Time series evaluation (2010-2014), divided by site",
    x = "Date",
    y = expression("GPP (gC m"^-2*" day"^-1*")"),
    color = "Data source:"
  ) +
  theme_minimal()

print(plot_combined_comparison_subset)
```

The time-series plots allow us to more closely analyze the cross-site model performance. Looking at the evaluation on the Laegern test set, we observe a systematic bias of the Davos-trained model. This model fails to capture the high summer GPP peaks, rarely exceeding a value of 8-10 gC m⁻² day⁻¹, while the observed GPP at Laegern often reaches over 15. This strong under-prediction bias is a result of the model training experience; because it has only learned from the colder, high-altitude Davos environment with its lower range of GPP, it is simply unable to predict the much higher productivity of the warmer, temperate Laegern site during its peak growing season, which is typical of a mixed deciduous-dominated forest with a seasonal leaf course.

Conversely, the time-series plot for the Davos test set shows that the Laegern-trained model, while performing reasonably well, tends to slightly over-predict GPP during the summer peak. The Laegern model, having been trained on an ecosystem with a warmer climate and prominently composed of deciduous forest, anticipates a level of summer productivity that the more constrained subalpine site cannot always reach.

The time-series plot of the combined test set confirms the very good adaptability of the model trained on both datasets. In the panel for the CH-Dav data, the combined model's predictions track the observed GPP almost perfectly, very similarly to the performance of the specialized Davos-only model. Even more interestingly, looking at the CH-Lae panel, the combined model successfully predicts the high summer GPP peaks, demonstrating that it has learned to produce higher GPP values when faced with Laegern-like conditions. This adaptive behavior, where the model is able to adjust its prediction range based on the input conditions of each site, shows that the combined model has not simply averaged the two sites behaviors, but has learned a more complex and generalizable function that integrates the distinct patterns from both ecosystems.


### 4.2. Limitations of the combined model and true out-of-sample considerations
When we train and test the combined model using data from both sites, it is an out-of-sample test for those specific data points, but not truly out-of-sample for the types of environments it has learned. The model is tuned to the characteristics of Jura montains and subalpine forests.

If this model was tested against data from a completely different environment, such as a Mediterranean forest in Spain, we would expect its performance to be poor. A Spanish site would likely have a much warmer and drier climate and completely different vegetation types, representing predictor values and environmental conditions far outside the domain of the training data. This is an extrapolation problem, the prediction would exceed the range of the training data. This shows that even a good model is only reliable for the specific types of environments it has been trained on. We cannot guarantee it will perform well in a completely new place it has never seen before. To make a model that can accurately predict for new climate zones, we must include data from those zones in its training.


## 5. Sources
Etzold, S., Ruehr, N. K., Zweifel, R., Dobbertin, M., Zingg, A., Pluess, P., ... & Buchmann, N. (2011). The carbon balance of two contrasting mountain forest ecosystems in Switzerland: similar annual trends, but seasonal differences. Ecosystems, 14, 1289-1309. DOI: 10.1007/s10021-011-9481-3.

Zielis, S., Etzold, S., Zweifel, R., Eugster, W., Haeni, M., & Buchmann, N. (2014). NEP of a Swiss subalpine forest is significantly driven not only by current but also by previous year's weather. Biogeosciences, 11(6), 1627-1635. DOI: 10.5194/bg-11-1627-2014.
