---
title: "Report 11.5"
author: "Francesco Lucano"
date: "03/06/2025" 
output:
  html_document:
    toc: true
---
## 1. Informations on site characteristics 
Here we provide information about the characteristics of the 2 study sites,  Lägeren (CH-Lae) and Davos Seehornwald (CH-Dav). The information are draw from Etzold et al., 2011 and Zielis et al., 2014. 

**Comparative characteristics of CH-Lae (Lägeren) and CH-Dav (Davos Seehornwald) forest sites**

---

**Location**

* CH-Lae: Located in the Swiss Jura mountain range, representative of the montane zone.
* CH-Dav: Situated in the Eastern Swiss Alps, representative of the subalpine zone.

---

**Elevation**

* CH-Lae: 682 
* CH-Dav: 1,639 

---

**Climate (mean annual values)**

**Air Temperature (20-year average 1989-2009)**

* CH-Lae: 7.4 °C 
* CH-Dav: 3.4 °C 

**Precipitation (20-year average 1989-2009)**

* CH-Lae: 1000 mm 
* CH-Dav: 1000 mm 

---

**Vegetation Type & Dominant Species**

* CH-Lae: A mixed deciduous-dominated forest with a seasonal leaf course. Key species include *Fagus sylvatica L.* (European beech), *Picea abies (L.) Karst* (Norway spruce), *Fraxinus excelsior L.* (ash), and *Acer pseudoplatanus L.*.
* CH-Dav: A coniferous forest, specifically a subalpine spruce forest, it's an evergreen site. Highly dominated by *Picea abies (L.) Karst* (Norway spruce).

---

**Maximum Leaf Area Index (LAI)**

* CH-Lae: 1.7-5.5 m$^2$ m$^{-2}$ (reflecting seasonal variation).
* CH-Dav: 3.9 m$^2$ m$^{-2}$ (more stable year-round).

---

**Summary**

In essence, the two sites show marked difference that will affect Gross Primary Production (GPP) and they will works well to assess model perfomance.

CH-Dav, represents a high-altitude subalpine environment characterized by a cold climate and snow precipitations that influence the GPP. Conversely, CH-Lae, a lower-altitude site, experiences a milder temperate climate. This altitudinal and climatic disparity directly translates into divergent growing season lengths and conditions.

This translates into very different vegetation types. CH-Dav is an established evergreen coniferous forest, predominantly composed of spruce, which sustains its canopy throughout the year. Conversely, CH-Lae is a younger mixed forest, prominently composed of deciduous trees, notably beech. This structural difference results in distinct seasonal changes in leaf area and overall canopy coverage at CH-Lae.


## 2. Data preparation
```{r setup, warning=FALSE, message=FALSE}
library(readr)     
library(dplyr)     
library(lubridate) 
library(ggplot2)   
library(rsample)   
library(recipes)   
library(caret)     
library(here)
library(yardstick)
library(tidyr)

# Load CH-Dav data
flux_davos_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Dav_FLUXNET2015_FULLSET_DD_1997-2014_1-3.csv"),
  delim = ","
)

# Load CH-Lae data
flux_laegeren_raw <- readr::read_delim(
  here::here("data", "FLX_CH-Lae_FLUXNET2015_FULLSET_DD_2004-2014_1-4.csv"),
  delim = ","
)
```

```{r, warning=FALSE, message=FALSE}
# We define a function in order to clean both dataset
clean_flux_data <- function(raw_df) { 
  df_cleaned <- raw_df %>%
    # We select only the columns we need
    select(
      TIMESTAMP, GPP_NT_VUT_REF, NEE_VUT_REF_QC,
      TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F
    ) %>%
    # We rename GPP column for simplicity
    rename(GPP = GPP_NT_VUT_REF) %>%
    # We convert TIMESTAMP to date object after making it a character 
    mutate(TIMESTAMP = as.Date(as.character(TIMESTAMP), format = "%Y%m%d"),
           # We replace -9999 values with NA
           across(where(is.numeric), ~na_if(.x, -9999))) %>%
    # We filter rows only to keep good quality data (value 1)
    filter(NEE_VUT_REF_QC == 1) %>% 
    # We remove rows where any of the remaining columns have NA values
    select(TIMESTAMP, GPP, TA_F, SW_IN_F, LW_IN_F, VPD_F, P_F, WS_F) %>%
    filter(complete.cases(.))
  return(df_cleaned)
}

# Then we apply it:
flux_davos_cleaned <- clean_flux_data(flux_davos_raw)
flux_laegeren_cleaned <- clean_flux_data(flux_laegeren_raw)
```

```{r, warning=FALSE, message=FALSE}
# We add a site_id column to each cleaned dataset
flux_davos_cleaned_labelled <- flux_davos_cleaned %>%
  mutate(site_id = "CH-Dav")

flux_laegeren_cleaned_labelled <- flux_laegeren_cleaned %>%
  mutate(site_id = "CH-Lae")

# We combine the two datasets
flux_combined_cleaned <- bind_rows(flux_davos_cleaned_labelled, flux_laegeren_cleaned_labelled)

# We convert site_id to a factor
flux_combined_cleaned <- flux_combined_cleaned %>%
  mutate(site_id = as.factor(site_id))

# Summary
print("Combined cleaned data head:")
print(head(flux_combined_cleaned))
print(dim(flux_combined_cleaned))
print("Summary of combined data:")
print(summary(flux_combined_cleaned))
print("Counts per site:")
print(table(flux_combined_cleaned$site_id))
```


## 3. Model training strategy


