---
title: "Report exercise 9.6"
author: "Francesco Lucano"
date: "`r Sys.Date()`"
output: html_document
---


### Setup
```{r setup, warning=FALSE, message=FALSE}
library(here)
library(ggplot2)
library(dplyr)  
library(readr)    

# Read the CSV file
df <- readr::read_delim(
  here("data", "df_for_stepwise_regression.csv"),
  delim = ","
)
```

```{r, warning=FALSE, message=FALSE}
# We define the target variable name
target_variable <- "GPP_NT_VUT_REF"

# We define potential predictor variable names, we exclude identifiers, timestamp, and the target itself
potential_predictors <- setdiff(names(df), c("siteid", "TIMESTAMP", "GPP_NT_VUT_REF"))

# We keep only numeric predictors 
numeric_predictors <- names(which(sapply(df[, potential_predictors], is.numeric)))
potential_predictors <- numeric_predictors 

# We create a subset of the dataframe with only the target and potential predictors
df_subset <- df[, c(target_variable, potential_predictors)]
```

```{r, warning=FALSE, message=FALSE}
# We remove any row that has at least a NA
df_clean <- na.omit(df_subset)

# Check dimensions after removing NAs
print(paste("Original dimensions:", dim(df)[1], "rows"))
print(paste("Dimensions after NA removal:", dim(df_clean)[1], "rows"))
```


### Bivariate model evaluation
```{r, warning=FALSE, message=FALSE}
# We calculate AIC of the intercept-only model
intercept_formula <- as.formula(paste(target_variable, "~ 1"))  # We create a formula with only the intercept 
intercept_model <- lm(intercept_formula, data = df_clean)       # We fit the intercept only model
intercept_aic <- AIC(intercept_model)                           # We calculate the AIC for the intercept-only model
print(paste("Intercept-only model AIC:", round(intercept_aic, 2)))

# We create an empty data frame to store results for bivariate models with two columns (predictor, aic)
bivariate_results <- data.frame(predictor = character(), aic = numeric(), stringsAsFactors = FALSE)

# We loop through each potential predictor to evaluate it alone
for (predictor in potential_predictors) {
  formula_str <- paste(target_variable, "~", predictor)
  current_formula <- as.formula(formula_str)
  # We try to fit the model, using tryCatch to handle potential errors (like collinearity)
  current_model <- tryCatch({ lm(current_formula, data = df_clean) }, error = function(e) { NULL })
  # Store AIC if model fitted successfully
  if (!is.null(current_model)) {    #Checks if model fitting was successful (if tryCatch didn't return NULL)
    current_aic <- AIC(current_model)
    bivariate_results <- rbind(bivariate_results, data.frame(predictor = predictor, aic = current_aic))
  }
}

# We find the best single predictor (lowest AIC)
best_bivariate_model <- NULL 
if (nrow(bivariate_results) > 0) {
  best_bivariate_model <- bivariate_results %>% arrange(aic) %>% slice(1)             # We arrange the data frame by the aic column in ascending order and we select the first row (lowest AI)
  print(paste("Best single predictor found:", best_bivariate_model$predictor))
  print(paste("AIC with this single predictor:", round(best_bivariate_model$aic, 2)))
}
```


### Stepwise forward regression algorithm implementation
```{r, warning=FALSE, message=FALSE}
selected_predictors <- c(best_bivariate_model$predictor)                    # selected_predictors holds the predictors in the current best model
remaining_predictors <- setdiff(potential_predictors, selected_predictors)  # remaining_predictors holds the predictors still left to try
best_model_aic <- best_bivariate_model$aic                                  # best_model_aic holds the AIC of our current best model

while (length(remaining_predictors) > 0) {      # The loop continues as long there are still left potential predictors
  
  best_aic_this_iteration <- Inf
  predictor_to_add_this_iteration <- NULL
  
  ############### We try adding one predictor (inner loop), 1 predictor per round ###############
  for (predictor in remaining_predictors) { 

    # We make a temporary model to test each predictor
    current_predictors_try <- c(selected_predictors, predictor)
    formula_str <- paste(target_variable, "~", paste(current_predictors_try, collapse = " + "))
    current_formula <- as.formula(formula_str)

    # We try to fit the temporary model, using tryCatch to handle potential errors (like collinearity)
    current_model <- tryCatch({ lm(current_formula, data = df_clean) }, error = function(e) { NULL })

    if (!is.null(current_model)) {          # We evaluate the model with AIC
        current_aic <- AIC(current_model)

        # We compare the AIC of the temporary model with the best AIC so far in this iteration
        if (current_aic < best_aic_this_iteration) {
            best_aic_this_iteration <- current_aic        # Retains the new AIC if it is smaller for this iteration
            predictor_to_add_this_iteration <- predictor  # Store the predictor associated with it
        }
    }
  }  # end of the inner loop

  ###################### Comparison with overall best model #################################
  # We check if the best predictor found in this round improves the overall best model AIC
  if (best_aic_this_iteration < best_model_aic) {
    # Yes, update our best model
    best_model_aic <- best_aic_this_iteration # Update overall best AIC
    selected_predictors <- c(selected_predictors, predictor_to_add_this_iteration) # Add the predictor
    remaining_predictors <- setdiff(remaining_predictors, predictor_to_add_this_iteration) # Remove predictor already tested
    # The while loop starts again

  } else {
    # No, adding the best remaining predictor didn't improve the AIC
    break # Stop the whole process and we exit the while loop
  }

} # End of the main while loop

######################################## Final results #################################
cat("\n----- Stepwise forward selection finished -----\n")
best_model_formula <- as.formula(paste(target_variable, "~", paste(selected_predictors, collapse = " + ")))
print(paste("Selected predictors:", paste(selected_predictors, collapse = ", ")))
print(paste("Final model formula:", deparse(best_model_formula)))
print(paste("Final model AIC:", round(best_model_aic, 2))) # This AIC comes from the stepwise process
```


### Final model fitting and result visualization
```{r, warning=FALSE, message=FALSE} 
# We fit the final linear model using the formula decided by the stepwise process
final_model <- lm(best_model_formula, data = df_clean)

print(summary(final_model)) 

# We add predictions as a new column to the dataframe and we use the final model to calculate the predicted GPP values
df_clean$predicted_GPP <- predict(final_model, newdata = df_clean)

# We create the scatter plot
plot_gpp <- ggplot(df_clean, aes(x = .data[[target_variable]], y = predicted_GPP)) +
  geom_point(alpha = 0.5) + 
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red", size = 1) + 
  labs(
    title = "Observed vs. Predicted GPP",
    subtitle = paste("Stepwise forward regression model (Final AIC =", round(best_model_aic, 2), ")"),
    x = "Observed GPP (GPP_NT_VUT_REF, umol CO2 / m^2 / s)", 
    y = "Predicted GPP (umol CO2 / m^2 / s)"
  ) +
  theme_minimal()

print(plot_gpp) 

################ Calculate Final RMSE ###########################
# Calculate residuals (actual - predicted)
residuals_final <- df_clean[[target_variable]] - df_clean$predicted_GPP
# Calculate RMSE
rmse_final <- sqrt(mean(residuals_final^2))
print(paste("Final model root mean squared error (RMSE):", round(rmse_final, 3)))
```


### Results and discussion
The aim of this work was to model the gross primary productivity GPP (GPP_NT_VUT_REF) using stepwise forward regression. We started the process with a simple intercept-only model, which had an AIC of 49959.43. The stepwise forward selection algorithm identified the following 10 predictors as the best set according to AIC: `PPFD_IN`, `LW_IN_F`, `VPD_F`, `TA_F_MDS`, `WS_F`, `USTAR`, `SW_IN_F`, `P_F`, `CO2_F_MDS`, and `PA_F`. The final selected model, including these 10 predictors, resulted in a lower AIC of 42838.15, meaning we improved the model.

The scatter plot shows that the correlation is positive as expected but there is quite a bit of scatter around the ideal prediction, especially at higher values where we observe an under-prediction of the model. Average values are more precise but there is a tendency to over-prediction, the lowest values are the most precise ones. 

Looking at the key numbers from the model summary, we see that  all 10 selected predictors were highly statistically significant (p < 0.001) in the final model. This suggests that they each contribute uniquely to the model. We also observe a moderate overall fit (Adjusted R-squared) of 0.5285, meaning that approximately 52.8% of the variance in GPP is explained by this model. The RMSE value is 2.307. The overall model significance is very high, the tiny  p-value (< 2.2e-16 ) indicates that the entire model is statistically better than just predicting the average GPP.

From an ecological perspective, the model's selection of predictors makes sense, including key drivers like light availability (`PPFD_IN`, `SW_IN_F`) and climatic factors (`TA_F_MDS`, `VPD_F`) which influence photosynthesis. The stepwise selection process also handled variable redundancy; for example, `TA_F` was likely excluded because the correlated `TA_F_MDS` variable was selected. This demonstrates how AIC is favorising a simpler model over a more complex one.  